\documentclass[11pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{enumitem}
\usepackage[hidelinks]{hyperref}

\geometry{margin=1in}
\setlist{nosep}
\setlength{\parskip}{0.55em}
\setlength{\parindent}{0pt}

\definecolor{codebackground}{RGB}{248,248,248}
\definecolor{codecomment}{RGB}{90,110,90}
\definecolor{codekeyword}{RGB}{20,70,160}
\definecolor{codestring}{RGB}{150,60,40}

\lstdefinestyle{pythonstyle}{
  language=Python,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{codekeyword}\bfseries,
  commentstyle=\color{codecomment},
  stringstyle=\color{codestring},
  backgroundcolor=\color{codebackground},
  frame=single,
  rulecolor=\color{black!15},
  breaklines=true,
  breakatwhitespace=true,
  showstringspaces=false,
  columns=fullflexible,
  keepspaces=true,
  tabsize=4
}

\lstdefinestyle{textstyle}{
  basicstyle=\ttfamily\small,
  backgroundcolor=\color{codebackground},
  frame=single,
  rulecolor=\color{black!15},
  breaklines=true,
  columns=fullflexible,
  keepspaces=true
}

\title{From Borges to Meaning: Progressive Reduction of the Library of Babel}
\author{}
\date{}

\begin{document}

\maketitle
\tableofcontents
\newpage

\phantomsection
\section*{Abstract}
\addcontentsline{toc}{section}{Abstract}

This teaser summarizes a sequence of mathematical transformations of \emph{The Library of Babel}. The historical Borges model is kept as a reference, but the progressive model chain starts with a normalized English-language Stage 0 so that every later comparison uses the same symbol inventory.

The shared rule is simple: the model may change the unit of generation, but it must not silently change the alphabet. At token level, all stages use the same punctuation pool:

\[
P = \{\mathtt{.},\ \mathtt{?},\ \mathtt{,},\ \mathtt{!}\}, \qquad |P| = 4.
\]

The common token alphabet is therefore:

\begin{lstlisting}[style=textstyle]
word . ? , !
\end{lstlisting}

The sequence moves from character-level entropy to lexical, syntactic, grammatical, semantic, topical, and eventually neural/reversible constraints.

\section{The Original Borges Combinatoric Baseline Model}

Borges defines books with:

\begin{itemize}
  \item 410 pages;
  \item 40 lines per page;
  \item 80 characters per line;
  \item an alphabet of 25 symbols.
\end{itemize}

Total character positions:

\[
C = 410 \times 40 \times 80 = 1{,}312{,}000.
\]

Historical Borges state space:

\[
25^{1{,}312{,}000}
\approx
1.95 \times 10^{1{,}834{,}097}.
\]

This model is the historical reference, but it is not the comparison baseline for the English token models because it uses a different alphabet.

\section{Stage 0: The English-Language Borges Baseline}

Stage 0 preserves Borges' physical book geometry but normalizes the character alphabet to the English symbols used by every later model:

\[
A_{\mathrm{en}} = \{a,b,\ldots,z,\ \text{space},\ \mathtt{.},\ \mathtt{?},\ \mathtt{,},\ \mathtt{!}\},
\qquad |A_{\mathrm{en}}| = 31.
\]

The number of independent character slots remains:

\[
C = 1{,}312{,}000.
\]

The normalized English baseline is therefore:

\[
31^{1{,}312{,}000}.
\]

In base-10 scale:

\[
\log_{10}(31^{1{,}312{,}000}) \approx 1{,}956{,}667.
\]

Stage 0 contains approximately:

\[
10^{1{,}956{,}667}
\]

possible books. All later reductions are measured against this shared English-symbol baseline.

\begin{lstlisting}[style=pythonstyle]
import hashlib

ENGLISH_CHARACTER_ALPHABET = list("abcdefghijklmnopqrstuvwxyz ") + [".", "?", ",", "!"]

def character_at(seed: str, position: int) -> str:
    data = f"{seed}:stage0:{position}".encode("utf-8")
    digest = hashlib.sha256(data).digest()
    value = int.from_bytes(digest[:8], "big")
    return ENGLISH_CHARACTER_ALPHABET[value % len(ENGLISH_CHARACTER_ALPHABET)]
\end{lstlisting}

\section{Stage 1: The Lexical Reduction Model}

Human language is not constructed from arbitrary character sequences. Stage 1 replaces character-level generation with token-level generation.

Assume:

\[
W = 100{,}000, \qquad P = 4.
\]

Using an average token width of approximately 6 characters, the token capacity of the same Borges-sized book is:

\[
N \approx \frac{1{,}312{,}000}{6} \approx 218{,}667.
\]

The lexical state space is:

\[
(W+P)^N = (100{,}000 + 4)^{218{,}667} = (100{,}004)^{218{,}667}.
\]

In base-10 scale:

\[
(100{,}004)^{218{,}667} \approx 10^{1{,}093{,}339}.
\]

\begin{lstlisting}[style=pythonstyle]
VOCABULARY_POOL = ["the", "house", "river", "memory", "darkness"]
PUNCTUATION_POOL = [".", "?", ",", "!"]
COMBINED_LEXICAL_POOL = VOCABULARY_POOL + PUNCTUATION_POOL

def lexical_token_at(seed: str, position: int) -> str:
    data = f"{seed}:stage1:{position}".encode("utf-8")
    digest = hashlib.sha256(data).digest()
    value = int.from_bytes(digest[:8], "big")
    return COMBINED_LEXICAL_POOL[value % len(COMBINED_LEXICAL_POOL)]
\end{lstlisting}

\section{Stage 2: The Syntactic Reduction Model (Punctuation Constraint)}

Stage 2 keeps the same vocabulary and punctuation pool, but forbids adjacent punctuation. This removes structurally invalid sequences such as:

\begin{lstlisting}[style=textstyle]
word . ? , !
\end{lstlisting}

If exactly $k$ punctuation marks appear in a sequence of $N$ tokens, valid non-adjacent placements are counted by:

\[
\binom{N-k+1}{k}.
\]

The total state space is:

\[
\sum_{k=0}^{\lfloor(N+1)/2\rfloor}
\binom{N-k+1}{k}
P^k
W^{N-k}.
\]

\begin{lstlisting}[style=pythonstyle]
def valid_sequence(tokens):
    punct = {".", "?", ",", "!"}

    for i in range(len(tokens) - 1):
        if tokens[i] in punct and tokens[i + 1] in punct:
            return False

    return True
\end{lstlisting}

\section{Stage 3: The Sentence-Structured Uniformity Constraint}

Stage 3 forces a fixed sentence-like block structure: exactly 15 words followed by exactly one punctuation token from the shared punctuation pool. Because comma is retained, the block should be understood as a sentence-or-clause block rather than only a final sentence terminator.

Pattern:

\begin{lstlisting}[style=textstyle]
W W W W W W W W W W W W W W W P
\end{lstlisting}

The number of complete blocks is:

\[
S = \left\lfloor \frac{N}{16} \right\rfloor.
\]

With $N \approx 218{,}667$, this gives $S \approx 13{,}666$. The state space is:

\[
W^{15S}P^S
\approx
(100{,}000)^{204{,}990} \times 4^{13{,}666}
\approx
10^{1{,}033{,}178}.
\]

\begin{lstlisting}[style=pythonstyle]
WORDS_PER_SENTENCE_BLOCK = 15
BLOCK_PUNCTUATION_POOL = [".", "?", ",", "!"]

def sentence_block(seed: str, block_index: int) -> str:
    words = [
        lexical_token_at(seed, block_index * 16 + i)
        for i in range(WORDS_PER_SENTENCE_BLOCK)
    ]

    data = f"{seed}:stage3:punct:{block_index}".encode("utf-8")
    value = int.from_bytes(hashlib.sha256(data).digest()[:8], "big")
    punct = BLOCK_PUNCTUATION_POOL[value % len(BLOCK_PUNCTUATION_POOL)]

    return " ".join(words) + punct
\end{lstlisting}

\section{Stage 4: The Categorical Context-Free Grammar Constraint}

Stage 4 partitions the vocabulary into grammatical categories and forces each block to follow a fixed syntactic template.

Canonical template:

\begin{lstlisting}[style=textstyle]
DET ADJ NOUN VERB DET NOUN PUNCT
\end{lstlisting}

For one block, the state space is:

\[
|D| \times |A| \times |N| \times |V| \times |D| \times |N| \times |P|.
\]

For $S$ blocks:

\[
(D \cdot A \cdot N \cdot V \cdot D \cdot N \cdot P)^S.
\]

\begin{lstlisting}[style=pythonstyle]
GRAMMAR_TEMPLATE_DEFINITION = [
    ("DET", ["the", "a", "every", "no"]),
    ("ADJ", ["dark", "ancient", "silent", "infinite"]),
    ("NOUN", ["river", "library", "mirror", "memory"]),
    ("VERB", ["contains", "reflects", "hides", "echoes"]),
    ("DET", ["the", "a", "every", "no"]),
    ("NOUN", ["page", "secret", "shadow", "symbol"]),
]
PUNCTUATION_POOL = [".", "?", ",", "!"]

def grammatical_sentence(seed: str, sentence_id: int) -> str:
    words = []

    for index, (_, category_words) in enumerate(GRAMMAR_TEMPLATE_DEFINITION):
        data = f"{seed}:stage4:{sentence_id}:{index}".encode("utf-8")
        value = int.from_bytes(hashlib.sha256(data).digest()[:8], "big")
        words.append(category_words[value % len(category_words)])

    data = f"{seed}:stage4:{sentence_id}:punct".encode("utf-8")
    value = int.from_bytes(hashlib.sha256(data).digest()[:8], "big")
    punct = PUNCTUATION_POOL[value % len(PUNCTUATION_POOL)]

    return " ".join(words) + punct
\end{lstlisting}

\section{Stage 5: The Markovian Semantic Adjacency Constraint}

Stage 5 replaces independent grammatical choices with a directed semantic graph. A valid book is a path through that graph.

Let $T$ be an adjacency matrix where $T_{ij}=1$ means token $j$ may follow token $i$. The approximate count of valid sequences is bounded asymptotically by:

\[
\lambda_{\max}^{N},
\]

where $\lambda_{\max}$ is the dominant eigenvalue of the transition matrix.

\begin{lstlisting}[style=pythonstyle]
SEMANTIC_ADJACENCY_GRAPH = {
    "the": ["ancient", "silent", "infinite", "library", "river"],
    "ancient": ["library", "river", "memory"],
    "library": ["contains", "hides", "reveals"],
    "river": ["reflects", "flows", "carries"],
    "contains": ["the", "every", "no"],
    "memory": [".", "?", ",", "!"],
}
\end{lstlisting}

\section{Stage 6: Topic-Coherent Manifolds via Latent Space Vectorization}

Stage 6 introduces a book-level topic variable. A coordinate chooses a topic vector $t$, and every generated token is biased toward words whose embeddings are close to that topic.

For a word embedding $v_i$, selection can be modeled as:

\[
P(w_i) \propto \exp\left(\beta \cdot \operatorname{cos\_sim}(v_i,t)\right).
\]

This stage adds long-range coherence: the whole book remains near a thematic manifold such as cosmology, grief, mathematics, theology, or war.

\section{Stage 7: Deterministic Neural Generative Steganography and Arithmetic Coding}

Stage 7 treats the coordinate as a deterministic random bitstream and uses a frozen language model as a probability estimator. Arithmetic decoding maps the coordinate into a sequence of tokens while respecting the model distribution:

\[
P(x_t \mid x_{<t}).
\]

The objective is to preserve naturalness while carrying coordinate entropy:

\[
\max H(Q) \quad \text{subject to} \quad D_{KL}(Q \parallel P) \leq \epsilon.
\]

This stage is theoretically powerful but requires strict determinism, quantized arithmetic, and a self-hosted model to keep coordinate-to-text mapping reproducible.

\section{Stage 8: Reversible Generative Flow Models}

Stage 8 is the most speculative endpoint: replace one-way autoregressive generation with reversible generative models. A coordinate maps into a latent vector, the flow maps the latent vector into text, and the inverse flow can verify or reconstruct the coordinate.

This is the mathematically cleanest version of a reversible semantic Library, but it is also the most resource-intensive and research-heavy stage.

\section{Final Observation}

The original Library is not terrifying because it contains all books. It is terrifying because meaningful books are drowned inside an ocean of entropy.

The coherent model chain is now:

\begin{lstlisting}[style=textstyle]
Historical Borges reference -> Stage 0 English baseline -> Stage 1 lexical -> Stage 2 punctuation-constrained -> Stage 3 sentence-block structured -> Stage 4 grammatical -> Stage 5 semantic graph -> Stage 6 topic manifold -> Stage 7 neural arithmetic coding -> Stage 8 reversible flows
\end{lstlisting}

Each stage decreases combinatoric volume, increases structural or semantic density, and moves the Library closer to the compressed region inhabited by human thought.

\end{document}
